Investigating Neolithic pottery usage in the Yanyuan Basin in Southwest China using organic residue analysis

### 2023-01-13 analysis/paper/paper.Rmd
worked on the method section

## Samples
We selected five ceramic fragments (Table XX) from Guijiabao for organic residue analysis, in which four samples were recovered from the trench 13 excavated in the 2017 field season and one sample was from the 2016 season. To further investigate the content and use of pottery vessels in an earlier period, we focus on potsherds excavated from the stratigraphic layer which dates to the early Neolithic period (5000-4500 BP.). During this period, the common vessel forms include jars, urns, bowls, and basins. For the pottery types, cord-marked grey pottery vessels with fine inclusions and cord-marked sand-tempered vessels in reddish-brown color with coarse inclusions are the two dominant types [@Hao2022]. Our samples are jar fragments from those two common types supposed to be used for food serving or storage according to their archaeological context that suggests living area [].

@@

We cleaned the surface of each potsherd to prevent external contamination from environmental factors and handling during excavation or post-excavation analysis. Approximately 3.5 grams powdered samples were extracted with a mixture of dichloromethane/methanol solution (9:1 v/v) using accelerated solvent extraction (ASE Dionex 350; at 100 degrees C). We conducted saponification by adding 10 mL 1M KOH in methanol and 1 mL water into the extracts and heated them in an oven at 70 degrees C overnight. Saponified extracts were acidified with 6M hydrochloric acid (HCl), and the acid fraction was extracted using hexane ($3\times10$ mL). A 5% aliquot of each sample was derivatized using 20 uL N,O-bis(trimethylsilyl)trifluoroacetamide (BSTFA) with 1% v/v trimethylchlorosilane (TMCS), and heated at 70 degrees C for one hour. After drying under a stream of nitrogen, the derivitized sample was dissolved in toluene (20 uL) prior to GC-MS.

The remaining 95% aliquot was fractioned by solid phase extraction (Aminopropyl, NH2, column with 0.5 g sorbent) with 8 mL dichloromethane:Isopropyl alcohol (3:1), 8 mL 4% acetic acid in ethyl ether, and 6 mL methanol respectively. The second fraction was methylated to produce fatty acid methyl esters (FAMEs) using 1 mL dry hexane and 2 mL 10% (v/v) acetyl chloride in dry methanol, and heated at 70 degree C overnight. The FAME derivatives were extracted with hexane ($3\times10$ mL) and blown down with nitrogen stream. About 5% aliquot dissolved in toluene was analyzed by GC-FID and the remaining sample was analyzed by GC-C-IRMS.

To further distinguish the source of fatty acids, we conducted stable carbon isotope analysis for those samples.

>>>>

## Samples
We selected five ceramic fragments from Guijiabao and one ceramic fragment from Gaoshan for organic residue analysis (Table XX). The four ceramic fragments were recovered from the trench 13 excavated in the 2017 field season and one sample was from the 2016 season. To understand the content and use of pottery vessels in the Yanyan region in an earlier period, we analyze potsherds from the stratigraphic layer which dates to the early Neolithic period (5000-4500 BP.) with potsherds from a later period (1000 BP.) for comparison. During the early Neolithic period, the common vessel forms include jars, urns, bowls, and basins. For the pottery types, cord-marked grey pottery vessels with fine inclusions and cord-marked sand-tempered vessels in reddish-brown color with coarse inclusions are the two dominant types [@Hao2022]. Our samples are jar fragments from those two common types that are supposed to be used for food serving or storage [].

@@

We cleaned the surface of each potsherd to prevent external contamination from environmental factors and handling during excavation or post-excavation analysis. Approximately 3.5 grams powdered samples were extracted with a mixture of dichloromethane/methanol solution (9:1 v/v) using accelerated solvent extraction (ASE Dionex 350; at 100 degrees C). We conducted saponification by adding 10 mL 1M KOH in methanol and 1 mL water into the extracts and heated them in an oven at 70 degrees C overnight. Saponified extracts were acidified with 6M hydrochloric acid (HCl), and the acid fraction was extracted using hexane ($3\times10$ mL). A 5% aliquot of each sample was derivatized using 20 uL N,O-bis(trimethylsilyl)trifluoroacetamide (BSTFA) with 1% v/v trimethylchlorosilane (TMCS), and heated at 70 degrees C for one hour. After drying under a stream of nitrogen, the derivitized sample was dissolved in toluene (20 uL) prior to GC-MS. Internal standards were added before and after the total lipid extraction.

The remaining 95% aliquot was fractioned by solid phase extraction (Aminopropyl, NH2, column with 0.5 g sorbent) with 8 mL dichloromethane:Isopropyl alcohol (3:1), 8 mL 4% acetic acid in ethyl ether, and 6 mL methanol respectively. The second fraction was methylated to produce fatty acid methyl esters (FAMEs) using 1 mL dry hexane and 2 mL 10% (v/v) acetyl chloride in dry methanol, and heated at 70 degree C overnight. The FAME derivatives were extracted with hexane ($3\times10$ mL) and blown down with nitrogen stream. About 5% aliquot dissolved in toluene was analyzed by GC-FID for quantification and the remaining was analyzed by GC-C-IRMS.

The GC-MS analysis was performed using an Agilent 7683 B series chromatograph attached to an Agilent 5975 inert mass selective detector. Samples were injected in splitless mode at 320 °C and compounds were separated with a VF-17ms column (60 m length, 0.25 mm i.d., 0.25 μm film thickness). Helium carrier is in constant flow rate of 1.24 ml/min. The temperature of the oven was started from 90 °C for 1 minute, and increased at a rate of 5 °C/min to reach 320 °C , where it was held for 23 min.
